---
title: "Untitled"
format: html
editor: visual
---


```{r}
library(readr)
library(readxl)
library(lubridate)
library(tidyverse)
library(dplyr)
library(stringi) # or library(stringr)
library(ggplot2)
```

* Load the data

```{r}
dir <- "/Users/johanna/Uni/ws_25_26/ecohydrology/ecohydro_project/data/Water stable isotope analysis"

fname_stand <- "+ 2024 cep in-house  standards.xlsx"
fname_data <- "20260309 Water stable isotope data.csv"

data <- read_csv(file.path(dir, fname_data))
standards <- read_excel(file.path(dir, fname_stand), skip = 1)
```

* Clean column names and data

```{r}
colnames(data) <- gsub(" ", "_", colnames(data))

# Correct time stamp
data$Time_Code <- ymd_hms(data$Time_Code)

cols <- data %>%
  select(where(is.character)) %>%
  colnames()

data$Identifier_2[which(data$Identifier_2==data$Identifier_2[55])] <- "spuelen"

# Remove blank spaces (using stringi for safety)
for(i in cols) {
  data[[i]] <- stringi::stri_replace_all_fixed(data[[i]], " ", "", vectorize_all = FALSE)
}

# Fix entries
data[data$Identifier_1=="SStandardheavy", "Identifier_1"] <- "Standardheavy"
data[data$Identifier_1=="Standardheavy2", "Identifier_1"] <- "Superheavy"

# Fix column names
data <- data %>%
  mutate(
    d.18_16.Mean = as.numeric(`d(18_16)Mean`),
    d.D_H.Mean = as.numeric(`d(D_H)Mean`)
  )

 head(data)
```

* Filter and summarize measurements

```{r}
# Only keep the last 3
ids <- unique(data$Identifier_1)[-c(1:3)] #removing standards
# Filter the data and keep only the last 3 injections per ID
data_short <- data %>%
  filter(Identifier_1 %in% ids) %>%
  group_by(Identifier_1) %>%
  arrange(Identifier_1, desc(Inj_Nr)) %>%  # Sort by ID and descending injection number
  slice_head(n = 3) %>%                     # Keep only the top 3 (last 3 injections)
  ungroup()

# Summarize
data_short <- data_short %>%
  group_by(Identifier_1) %>%
  summarise(
    d.18_16.Mean.Mean = mean(d.18_16.Mean, na.rm = TRUE),
    d.18_16.Mean.Sd = sd(d.18_16.Mean, na.rm = TRUE),
    d.D_H.Mean.Mean = mean(d.D_H.Mean, na.rm = TRUE),
    d.D_H.Mean.Sd = sd(d.D_H.Mean, na.rm = TRUE),
    Time_Code = first(Time_Code),
    .groups = "drop"
  )

 head(data_short)
```

* Filter measured standards

```{r}
ids_standards <- unique(data$Identifier_1)[1:3]
data_standards <- data %>%
  filter(Identifier_1 %in% ids_standards) %>%
  group_by(Port) %>%
  arrange(Port, desc(Inj_Nr)) %>%  # Sort by ID and descending injection number
  slice_head(n = 3) %>%            # Keep only the top 3 (last 3 injections)
  ungroup() %>%
  filter(
    !Port %in% c(
      "1-01", "1-02", "1-94", "1-93", # Keep only last of first (1-03) and last (1-95) block for standard medium
      "1-04", "1-05", "1-90", "1-91", # of Standardheavy (1-06 and 1-92)
      "1-07", "1-08", "1-87", "1-88"  # and of Superheavy (1-09 and 1-89)
    )
  )

# Identify jumps
test1 <- data_standards %>% filter(Identifier_1 == "Standardmedium")
plot(test1$Time_Code, test1$d.18_16.Mean)
test2 <- data_standards %>% filter(Identifier_1 == "Standardheavy")
plot(test2$Time_Code, test2$d.18_16.Mean)
test3 <- data_standards %>% filter(Identifier_1 == "Superheavy")
plot(test3$Time_Code, test3$d.18_16.Mean)
```

* Summarize measured standards

```{r}
# Summarize standards
data_standards_means <- data_standards %>%
  group_by(Port) %>%
  summarise(
    d.18_16.Mean.Mean = mean(d.18_16.Mean, na.rm = TRUE),
    d.18_16.Mean.Sd = sd(d.18_16.Mean, na.rm = TRUE),
    d.D_H.Mean.Mean = mean(d.D_H.Mean, na.rm = TRUE),
    d.D_H.Mean.Sd = sd(d.D_H.Mean, na.rm = TRUE),
    Identifier_1 = first(Identifier_1),
    Time_Code = first(Time_Code),
    .groups = "drop"
  )

test1 <- data_standards_means %>% filter(Identifier_1 == "Standardmedium")
plot(test1$Time_Code, test1$d.18_16.Mean.Mean)
test2 <- data_standards_means %>% filter(Identifier_1 == "Standardheavy")
plot(test2$Time_Code, test2$d.18_16.Mean.Mean)
test3 <- data_standards_means %>% filter(Identifier_1 == "Superheavy")
plot(test3$Time_Code, test3$d.18_16.Mean.Mean)

test1 <- data_standards_means %>% filter(Identifier_1 == "Standardmedium")
plot(test1$Time_Code, test1$d.D_H.Mean.Mean)
test2 <- data_standards_means %>% filter(Identifier_1 == "Standardheavy")
plot(test2$Time_Code, test2$d.D_H.Mean.Mean)
test3 <- data_standards_means %>% filter(Identifier_1 == "Superheavy")
plot(test3$Time_Code, test3$d.D_H.Mean.Mean)
```

* Fit model from the first standard block to calibarte standards

```{r}
# Fit model for calibration
Standardmedium_d.18_16_gt <- -9.05
Standardheavy_d.18_16_gt <- -4.91
Superheavy_d.18_16_gt <- -0.18
Standardmedium_d.D_H_gt <- -63.63
Standardheavy_d.D_H_gt <- -10.18
Superheavy_d.D_H_gt <- 54.05
standards <- data.frame(standard= c("Standardmedium", "Standardheavy", "Superheavy"),
                        d.18_16= c(Standardmedium_d.18_16_gt, Standardheavy_d.18_16_gt, Superheavy_d.18_16_gt),
                        d.D_H= c(Standardmedium_d.D_H_gt, Standardheavy_d.D_H_gt, Superheavy_d.D_H_gt))

# kalibriert alle laufenden Standards mit dem ersten Standardblock
first_block_calib_data <- full_join(
  data_standards_means[1:3,], # first block only
  standards,
  by= join_by(Identifier_1 == standard)
)

# -- 18O

# Compute coefficients
lm_d.18_16 <- lm(d.18_16 ~ d.18_16.Mean.Mean, first_block_calib_data)
data_short$d.18_16.corr <- unname(
  predict(lm_d.18_16, newdata = data.frame("d.18_16.Mean.Mean"= data_short$d.18_16.Mean.Mean))
)

# Plot calibration
plot(d.18_16~d.18_16.Mean.Mean, first_block_calib_data, col=as.factor(Identifier_1), pch=16,
     xlab="measured Delta1816", ylab="Standard Delta1816")
curve(lm_d.18_16$coefficients[2]*x+ lm_d.18_16$coefficients[1], add=T)

# -- 2H

# Compute coefficients
lm_d.D_H <- lm(d.D_H ~ d.D_H.Mean.Mean, first_block_calib_data)
data_short$d.D_H.corr <- unname(
  predict(lm_d.D_H, newdata = data.frame("d.D_H.Mean.Mean"= data_short$d.D_H.Mean.Mean))
)

# Plot calibration
plot(d.D_H~d.D_H.Mean.Mean, first_block_calib_data, col=as.factor(Identifier_1), pch=16,
     xlab="Measured Delta D [‰]", ylab="Standard Delta D [‰]", fontsize = 16)
curve(lm_d.D_H$coefficients[2]*x+ lm_d.D_H$coefficients[1], add=T)
```

## Drift correction for 18O

* Compute difference of calibrated standards to ground truth standards for 18O

```{r}

# -- 18O

# Calibrate standards
slope <- lm_d.18_16$coefficients[2]
intercept <- lm_d.18_16$coefficients[1]
data_standards_means$d.18_16.calib <- intercept + slope * data_standards_means[, "d.18_16.Mean.Mean"]

# berechnet dann die Differenz der Werte zu den Soll Werten (18O Dif und 2 H Dif)
n <- nrow(data_standards_means)
data_standards_diff <- data_standards_means %>%
  mutate(d.18_16.diff = case_when(
      Identifier_1 == "Standardmedium" ~ Standardmedium_d.18_16_gt - d.18_16.Mean.Mean,
      Identifier_1 == "Standardheavy" ~ Standardheavy_d.18_16_gt - d.18_16.Mean.Mean,
      Identifier_1 == "Superheavy" ~ Superheavy_d.18_16_gt - d.18_16.Mean.Mean
    )
  )

# Dann könnt ihr diese Differenz gegen die Injektionnummer oder besser Zeit bei euch auftragen sehen, wie die Werte wegdriften.
plot(data_standards_diff$Time_Code, data_standards_diff$d.18_16.diff)
```

* Fit a regression model for the drift

```{r}
# --- Drift regression model

# Wenn das so aussieht, wie bei 2H in dem Beispiel, dann könnt ihr für x die Zeit/Injektionsnummer einsetzen und
# diesen Wert dann auf euren standardisiertes Ergebnis draufrechnen (Tab Summary auf der rechten Seite bei 18Ocor und 2Hcor).

# convert time stamp -> to time past since first measurmenet
data_standards_diff <- head(data_standards_diff, nrow(data_standards_diff) - 3)
data_standards_diff$Time_Numeric <- as.numeric(difftime(data_standards_diff$Time_Code,
                                                       min(data_standards_diff$Time_Code),
                                                       units = "hours"))

# Fit the model
lm_d.18_16_drift <- lm(d.18_16.diff ~ Time_Numeric + I(Time_Numeric^2), data_standards_diff)
summary(lm_d.18_16_drift)

# - Regression plot
new_time <- seq(min(data_standards_diff$Time_Numeric),
                max(data_standards_diff$Time_Numeric),
                length.out = 100)
pred <- predict(lm_d.18_16_drift,
                newdata = data.frame(Time_Numeric = new_time),
                se.fit = TRUE)
pred_df <- data.frame(
  Time_Numeric = new_time,
  fit = pred$fit,
  se.fit = pred$se.fit,
  lower = pred$fit - 1.96 * pred$se.fit,
  upper = pred$fit + 1.96 * pred$se.fit
)
ggplot() +
  geom_point(data = data_standards_diff,
             aes(x = Time_Numeric, y = d.18_16.diff)) +
  geom_line(data = pred_df,
            aes(x = Time_Numeric, y = fit), color = "red") +
  geom_ribbon(data = pred_df,
              aes(x = Time_Numeric, ymin = lower, ymax = upper),
              alpha = 0.2, fill = "blue") +
  labs(x = "Time (hours since first observation)",
       y = "Delta1816 Difference to ground truth") +
  theme_minimal()
```

* Predict the drift from time stamps and plot corrected vs original data

```{r}
# --- Predict drift

# Timestamp to numeric
data_standards_means$Time_Numeric <- as.numeric(difftime(data_standards_means$Time_Code,
                                                         min(data_standards_means$Time_Code),
                                                         units = "hours"))
# Get coeffificents and intercept for drift prediction
slope_drift_lin <- lm_d.18_16_drift$coefficients[2]
slope_drift_quad <- lm_d.18_16_drift$coefficients[3]
intercept_drift <- lm_d.18_16_drift$coefficients[1]
data_standards_means <- data_standards_means %>%
  mutate(
    d.18_16.drift = (
      intercept_drift + slope_drift_lin * data_standards_means$Time_Numeric +
      slope_drift_quad * (data_standards_means$Time_Numeric^2)
    )
  )

# - Correct the drift by adding the corrected difference
correction <- (data_standards_means$d.18_16.calib + data_standards_means$d.18_16.drift)
data_standards_means$d.18_16.corr <- correction$d.18_16.Mean.Mean

# --- Plot corrected standardised vs original data

# Reshape the data
plot_data <- data_standards_means %>%
  filter(Identifier_1 %in% c("Standardmedium", "Standardheavy", "Superheavy")) %>%
  pivot_longer(
    cols = c(d.18_16.Mean.Mean, d.18_16.corr),
    names_to = "Value_Type",
    values_to = "Value"
  )

ggplot(plot_data, aes(x = Time_Code, y = Value, color = Value_Type)) +
  geom_point() +
  geom_line() +
  facet_wrap(~ Identifier_1, scales = "free_y", ncol = 1) +
  labs(x = "Time", y = "Value", color = "Value Type") +
  theme_minimal() +
  scale_color_manual(values = c("black", "red"))
```

## Drift correction for 2H

* Compute difference of calibrated standards to ground truth standards for 2H

```{r}

# -- 2H

# Calibrate standards
slope <- lm_d.D_H$coefficients[2]
intercept <- lm_d.D_H$coefficients[1]
data_standards_means$d.D_H.calib <- intercept + slope * data_standards_means[, "d.D_H.Mean.Mean"]

# berechnet dann die Differenz der Werte zu den Soll Werten (18O Dif und 2 H Dif)
n <- nrow(data_standards_means)
data_standards_diff <- data_standards_diff %>%
  mutate(d.D_H.diff = case_when(
      Identifier_1 == "Standardmedium" ~ Standardmedium_d.D_H_gt - d.D_H.Mean.Mean,
      Identifier_1 == "Standardheavy" ~ Standardheavy_d.D_H_gt - d.D_H.Mean.Mean,
      Identifier_1 == "Superheavy" ~ Superheavy_d.D_H_gt - d.D_H.Mean.Mean
    )
  )

# Dann könnt ihr diese Differenz gegen die Injektionnummer oder besser Zeit bei euch auftragen sehen, wie die Werte wegdriften.
plot(data_standards_diff$Time_Code, data_standards_diff$d.D_H.diff)
```

* Fit a regression model for the drift

```{r}
# --- Drift regression model

# Wenn das so aussieht, wie bei 2H in dem Beispiel, dann könnt ihr für x die Zeit/Injektionsnummer einsetzen und
# diesen Wert dann auf euren standardisiertes Ergebnis draufrechnen (Tab Summary auf der rechten Seite bei 18Ocor und 2Hcor).

# Fit the model
lm_d.D_H_drift <- lm(d.D_H.diff ~ Time_Numeric + I(Time_Numeric^2), data_standards_diff)
summary(lm_d.D_H_drift)

# - Regression plot
new_time <- seq(min(data_standards_diff$Time_Numeric),
                max(data_standards_diff$Time_Numeric),
                length.out = 100)
pred <- predict(lm_d.D_H_drift,
                newdata = data.frame(Time_Numeric = new_time),
                se.fit = TRUE)
pred_df <- data.frame(
  Time_Numeric = new_time,
  fit = pred$fit,
  se.fit = pred$se.fit,
  lower = pred$fit - 1.96 * pred$se.fit,
  upper = pred$fit + 1.96 * pred$se.fit
)

p <- ggplot() +
  geom_point(data = data_standards_diff,
             aes(x = Time_Numeric, y = d.D_H.diff), color = "grey") +
  geom_line(data = pred_df,
            aes(x = Time_Numeric, y = fit), color = "darkgrey") +
  geom_ribbon(data = pred_df,
              aes(x = Time_Numeric, ymin = lower, ymax = upper),
              alpha = 0.2, fill = "grey") +
  labs(x = "Time t (hours since first observation)",
       y = "Delta D ground truth - measurement [‰] ") +
  theme_minimal()

# Save the plot using ggsave
ggsave(
  filename = file.path(outdir, "2H_regression.png"),
  plot = p,
  width = 5,
  height = 4,
  dpi = 300,  # Optional: Set DPI for higher quality
  bg = "white"
)
```

* Predict the drift from time stamps and plot corrected vs original data

```{r}
# --- Predict drift

# Get coeffificents and intercept for drift prediction
slope_drift_lin.D_H <- lm_d.D_H_drift$coefficients[2]
slope_drift_quad.D_H <- lm_d.D_H_drift$coefficients[3]
intercept_drift.D_H <- lm_d.D_H_drift$coefficients[1]
data_standards_means <- data_standards_means %>%
  mutate(
    d.D_H.drift = (
      intercept_drift.D_H + slope_drift_lin.D_H * data_standards_means$Time_Numeric +
      slope_drift_quad.D_H * (data_standards_means$Time_Numeric^2)
    )
  )

# - Correct the drift by adding the corrected difference
correction <- (data_standards_means$d.D_H.calib + data_standards_means$d.D_H.drift)
data_standards_means$d.D_H.corr <- correction$d.D_H.Mean.Mean

# --- Plot corrected standardised vs original data

# Reshape the data
plot_data <- data_standards_means %>%
  filter(Identifier_1 %in% c("Standardmedium", "Standardheavy", "Superheavy")) %>%
  pivot_longer(
    cols = c(d.D_H.Mean.Mean, d.D_H.corr),
    names_to = "Value_Type",
    values_to = "Value"
  )

ggplot(plot_data, aes(x = Time_Code, y = Value, color = Value_Type)) +
  geom_point() +
  geom_line() +
  facet_wrap(~ Identifier_1, scales = "free_y", ncol = 1) +
  labs(x = "Time", y = "Value", color = "Value Type") +
  theme_minimal() +
  scale_color_manual(values = c("black", "red"))
#dev.off()
```

## Correct measurements

* Correct 18O

```{r}
data_short$Time_Numeric <- as.numeric(difftime(data_short$Time_Code,
                                               min(data_short$Time_Code),
                                                  units = "hours"))

data_short$drift.18_16 <- unname(predict(lm_d.18_16_drift, newdata = data.frame(Time_Numeric=data_short$Time_Numeric)))

data_short$d.18_16.corr.drift <- data_short$d.18_16.corr + data_short$drift.18_16

plot(d.18_16.corr~Time_Code, data= data_short)
points(d.18_16.corr.drift~Time_Code, data= data_short, col="orange")
```

* Correct 2H

```{r}
data_short$drift.D_H <- unname(predict(lm_d.D_H_drift, newdata = data.frame(Time_Numeric=data_short$Time_Numeric)))

data_short$d.D_H.corr.drift <- data_short$d.D_H.corr + data_short$drift.D_H

data_short_no_label <- data_short %>% filter(d.D_H.corr < 4000)

plot(d.D_H.corr ~ Time_Code, data= data_short_no_label)
points(d.D_H.corr.drift ~ Time_Code, data= data_short_no_label, col="orange")

plot(d.D_H.corr ~ Time_Code, data= data_short)
points(d.D_H.corr.drift ~ Time_Code, data= data_short, col="orange")
```
* Create final dataframe with corrected values

```{r}

final_df <- data_short

final_df <- final_df %>%
  mutate(
    species = case_when(
      substr(Identifier_1, 1, 2) == "SD" ~ "Douglas fir",
      substr(Identifier_1, 1, 2) == "SK" ~ "Pine",
      substr(Identifier_1, 1, 1) == "D"  ~ "Douglas fir",
      substr(Identifier_1, 1, 1) == "K"  ~ "Pine",
      TRUE ~ NA_character_
    )
  )

final_df$level <- ifelse(substr(final_df$Identifier_1, 4, 5)=="hi", "high",
                           ifelse(substr(final_df$Identifier_1, 4, 5)=="mi","mid",
                                  ifelse(substr(final_df$Identifier_1, 4, 5)=="lo", "low", NA)))
final_df$level <- factor(final_df$level, levels = c("low", "mid", "high"))

final_df$Date <- as.Date(ifelse(substr(final_df$Identifier_1, start= 8, 9) %in% c("1", "_1"), as.Date("2026-02-25"),
                    ifelse(substr(final_df$Identifier_1, start= 8, 9) %in% c("2", "_2"), as.Date("2026-02-26"),
                           ifelse(substr(final_df$Identifier_1, start= 8, 9) %in% c("3", "_3"), as.Date("2026-02-27"), NA))      ))

final_df$Date[which(final_df$Identifier_1 %in% c("D1_C", "D2_C", "D3_C", "K1_C", "K2_C", "K3_C"))] <- as.Date("2026-02-24")

final_df <- final_df %>% select(-Time_Numeric)

View(final_df)
```
* Save data

```{r}
write_csv(final_df, file.path(dir, "20260309_water_stable_isotope_data_corrected_v2.csv"))
```